---
title: "README"
author: "Zachary McCaw"
date: "`r Sys.Date()`"
output: 
  html_document: 
    keep_md: TRUE
--- 

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

# Description

This package provides functions for inference on the difference and ratio in the areas under mean cumulative count curves comparing two treatment arms. 

# Installation

```{R, eval=FALSE}
devtools::install_github(repo = 'zrmacc/MCC')
```

# Examples

Synthetic example data in the format expected by this package may be loaded via:

```{R}
library(MCC)
data(mcc_data)
head(mcc_data)
```

Here: 

* `idx` is the subject index. 
* `time` is the event time. 
* `status` is coded 1 for the recurrent event, and 0 otherwise.
* `arm` is the treatment arm, coded as 1 for treatment, 0 for reference. 

For analyses of other data sets, arm and status should have the same coding. 

## Difference and Ratio of AUCs

To find a confidence interval and p-vaue for the difference and ratio in areas under the mean cumulative count curve at time $\tau = 60$: 
```{r}
set.seed(100)
results <- CompareAUC(
  time = mcc_data$time,
  status = mcc_data$status,
  arm = mcc_data$arm,
  idx = mcc_data$idx,
  tau = 60,
  reps = 2000,
  alpha = 0.05
)
show(results)

```
