---
output:
  github_document
--- 

# Comparison of mean cumulative count curves via the area under the curve

[![R-CMD-check](https://github.com/zrmacc/MCC/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/zrmacc/MCC/actions/workflows/R-CMD-check.yaml)

Zachary R. McCaw <br>
Updated: 2026-06-25

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all()
} else {
  library(MCC)
}
library(dplyr)
```

## Installation

```{R, eval=FALSE}
remotes::install_github("zrmacc/MCC", build_vignettes = TRUE)
```

## Description

This package provides estimation and inference for the mean cumulative function (MCF) and the area under the MCF (AUMCF). The mean cumulative count (MCC) curve is an alias for the MCF. Estimation and inference for the MCF follow [Ghosh and Lin (2000)](https://onlinelibrary.wiley.com/doi/abs/10.1111/j.0006-341X.2000.00554.x); corresponding procedures for the AUMCF are described by [Gronsbell et al. (2026)](https://academic.oup.com/biometrics/article/82/2/ujag053/8689483).

## Univariate

Univariate analysis considers a single multiple- or recurrent-events process, subject to right-censoring and terminal events.

### Methods

#### MCF estimation

For each study arm, define $N(\cdot)$ as the counting process for events of interest (terminal and non-terminal), $Y(\cdot)$ as the number of subjects at risk, and $S(\cdot)$ as the probability of not having experienced a terminal event. The MCF $\mu(t)$ at time $t$ is estimated by:

$$
\mu(t) = \int_{0}^{t}\ \hat{S}(u)\ \frac{ dN(u) }{ Y(u) }
$$

Here $\hat{S}(u)$ is the Kaplan-Meier estimate of the probability of being terminal event-free; $dN(u)$ is the number of events of interest at time $u$; and $Y(u)$ is the number of subjects who remain at risk. The AUMCF to time $\tau$ is $\hat\alpha(\tau) = \int_0^\tau \hat\mu(u)\,du$.

#### Influence function

Let $\theta(\tau)$ denote the MCF or AUMCF at time $\tau$. The influence function $\psi_{i}(\tau)$ measures the contribution of subject $i$ to the estimation error:

$$
\hat{\theta}(\tau) - \theta(\tau) = \frac{1}{n}\sum_{i=1}^{n}\psi_{i}(\tau) + o_{p}(n^{-1/2})
$$

The influence function contributions estimate the variance of $\hat{\theta}(\tau)$:

$$
\hat{\mathbb{V}}(\hat{\theta}) = \frac{1}{n^2}\sum_{i=1}^{n}\psi_{i}^{2}(\tau).
$$

### Data

The function `GenData` simulates example data. Recurrent event times are generated from a Poisson process until censoring or death. Optionally, a gamma `frailty_variance` correlates patient-specific event and death rates. The example below has 100 patients per arm, `tau = 4`, and an 80% recurrent-event rate in the treatment arm relative to control.

```{R}
covariates <- data.frame(
  arm = c(rep(1, 100), rep(0, 100))
)
data <- MCC::GenData(
  beta_event = c(log(0.8)),
  covariates = covariates,
  frailty_variance = 0.2,
  tau = 4
)
head(data)
```

The essential columns are:

* `idx`, the subject index.
* `time`, the observation time.
* `status`, coded 0 for censoring, 1 for an event, 2 for death (or any competing terminal event).
* `arm`, coded as 1 for treatment, 0 for reference.

Each subject should experience an observation-terminating event (censoring or death). For other data sets, use the same arm and status coding.

#### Observation-terminating events

In the recurrent-events setting, a subject may remain at risk after an event of interest. An *observation-terminating* event (censoring or competing risk) removes the subject from the risk set. If a subject lacks such an event, they are assumed to remain at risk indefinitely. By default, `CompareAUCs` adds a censoring time immediately after the last event of interest. For example, if the data for subject 1 were:

```{r, echo = FALSE}
df <- data.frame(
  idx = c(1, 1, 1),
  time = c(2, 3, 5),
  status = c(1, 1, 1)
)
show(df)
```

then, for analysis, the subject is assumed censored after the last event:

```{r, echo = FALSE}
df <- data.frame(
  idx = c(1, 1, 1, 1),
  time = c(2, 3, 5, 5),
  status = c(1, 1, 1, 0)
)
show(df)
```

Set `cens_after_last = FALSE` if the subject should remain at risk indefinitely.

#### Terminal events of interest

When a fatal event is of interest, record it with two rows: `status = 1` (event occurred) and `status = 2` (event was terminal). For example, subject 1 had three events and the third was terminal:

```{r, echo = FALSE}
df <- data.frame(
  idx = c(1, 1, 1, 1),
  time = c(2, 3, 5, 5),
  status = c(1, 1, 1, 2)
)
show(df)
```

By contrast, subject 2 had three non-terminal events and is censored after the last by default:

```{r, echo = FALSE}
df <- data.frame(
  idx = c(2, 2, 2, 2),
  time = c(2, 3, 5, 5),
  status = c(1, 1, 1, 0)
)
show(df)
```

Censoring (`status = 0`) and terminal events (`status = 2`) both remove a subject from the risk set, but only censoring leaves open the possibility of future events.

### Analyses

#### Single-arm AUC

```{r}
auc <- MCC::SingleArmAUC(
  data %>% dplyr::filter(arm == 0),
  boot = TRUE,
  reps = 200,
  tau = 4
)
show(auc)
```

#### Two-arm comparison

```{r}
aucs <- MCC::CompareAUCs(
  data,
  tau = 4,
  boot = TRUE,
  perm = TRUE,
  reps = 200,
  alpha = 0.05
)
show(aucs)
```

* `tau` is the truncation time for the AUC.
* `boot` constructs bootstrap confidence intervals.
* `perm` performs permutation tests for the difference and ratio of AUCs.
* `reps` is the number of bootstrap/permutation replicates (grouped by `idx`, stratified by `strata` if applicable).
* `alpha` is 1 minus the desired CI coverage.

#### Jump-weighted analysis

The `jump_weights` argument scales the jump in the MCF at each recurrent event (`status == 1`). The example below weights each event by the patient's cumulative event count:

```{r, results = "hold"}
data <- data %>%
  dplyr::group_by(idx) %>%
  dplyr::mutate(jump_weights = dplyr::row_number()) %>%
  dplyr::ungroup()

cat("Jump weights for the first 10 records.\n")
data %>%
  dplyr::select(idx, time, status, jump_weights) %>%
  dplyr::slice(1:10)
```

```{r}
aucs <- MCC::CompareAUCs(
  data,
  tau = 4,
  alpha = 0.05,
  jump_weights = data$jump_weights
)
show(aucs)
```

#### Stratified analysis

The example below adds a binary stratification factor; the event rate in stratum 1 is increased by 20%.

```{r}
covariates <- data.frame(
  arm = c(rep(1, 100), rep(0, 100)),
  strata = stats::rbinom(200, 1, 0.25)
)
data <- MCC::GenData(
  beta_event = c(log(0.8), log(1.2)),
  covariates = covariates,
  frailty_variance = 0.2,
  tau = 4
)
aucs <- MCC::CompareAUCs(
  data,
  strata = data$strata,
  tau = 4,
  boot = TRUE,
  perm = TRUE,
  reps = 200,
  alpha = 0.05
)
show(aucs)
```

#### Outputs

The `CompareAUCs` object includes:

* `@StratumAreas`: stratum-specific AUCs per arm.
```{r}
aucs@StratumAreas
```

* `@MargAreas`: marginal AUCs per arm.
```{r}
aucs@MargAreas
```

* `@CIs`: confidence intervals for the difference and ratio of AUCs.
```{r}
aucs@CIs
```

* `@MCF`: per-arm MCFs.
```{r}
head(aucs@MCF)
```

* `@Pvals`: bootstrap and permutation p-values.
```{r}
aucs@Pvals
```

* `@Reps`: bootstrap and permutation test statistics.

#### Influence function

`InfluenceFunction` calculates per-subject influence contributions for the MCF (`type = "MCF"`) or AUMCF (`type = "AUC"`) at time $\tau$; see Methods for the definition of $\psi_i(\tau)$.

```{r}
psi_auc <- MCC::InfluenceFunction(
  data %>% dplyr::filter(arm == 0),
  tau = 4,
  type = "AUC"
)
head(psi_auc)
```

#### Adjusted AUCs

With continuous covariates, `CompareAUCs` uses an augmentation estimator. Do not supply both `strata` and `covar`; use `model.matrix` to combine them if needed.

```{r, results = "hold"}
set.seed(100)
n <- 1000
covariates <- data.frame(
  arm = c(rep(1, n), rep(0, n)),
  x1 = c(stats::rnorm(n, mean = -1), stats::rnorm(n, mean = 1)),
  x2 = c(stats::rnorm(n, mean = 1), stats::rnorm(n, mean = -1))
)
data <- MCC::GenData(
  beta_event = c(log(0.5), log(0.8), log(1.2)),
  covariates = covariates,
  base_death_rate = 0.25,
  base_event_rate = 1,
  frailty_variance = 0.2,
  tau = 4
)
unadj_aucs <- MCC::CompareAUCs(data, tau = 4, alpha = 0.05)
adj_aucs <- MCC::CompareAUCs(
  data,
  tau = 4,
  alpha = 0.05,
  covar = data %>% dplyr::select(x1, x2)
)
show(unadj_aucs)
show(adj_aucs)
```

#### Plotting

```{r, fig.width=7, fig.height=4}
q_uni <- MCC::PlotMCFs(
  data = data,
  tau = 4,
  x_breaks = seq(0, 4, by = 1),
  color_labs = c("Control", "Treatment")
)
q_uni
```

See the [plotting vignette](https://github.com/zrmacc/MCC/blob/master/vignettes/plotting.pdf) for additional options.

## Multivariate

Multivariate analysis considers several potentially dependent multiple- or recurrent-events processes, each subject to right-censoring and terminal events.

### Methods

For event types $k = 1,\ldots,K$, the arm–process estimand is the AUMCF $\alpha_{ak}(\tau)$. The multivariate contrast is

$$
\hat\Delta(\tau) = \bigl(\hat\Delta_1(\tau),\ldots,\hat\Delta_K(\tau)\bigr)^\top,
\qquad \hat\Delta_k(\tau) = \hat\alpha_{1k}(\tau) - \hat\alpha_{0k}(\tau).
$$

Optional `process_weights` $w \in \mathbb{R}^K$ defines a secondary scalar estimand $w^\top\hat\Delta(\tau)$, returned in `@Weighted`. Per-event jump weighting is not supported for multivariate data; each recurrent event contributes a jump of size 1.

Arm-specific influence vectors $\hat\phi_{ia} \in \mathbb{R}^K$ are stacked from per-process influence functions. The arm covariance is

$$
\hat\Sigma_a = \frac{1}{n_a} \sum_{i: A_i=a} \hat\phi_{ia}\hat\phi_{ia}^\top,
$$

and $\widehat{\mathrm{Cov}}(\hat\Delta) = \hat\Sigma_1/n_1 + \hat\Sigma_0/n_0$.

When patient $i$ is eligible for only a subset of processes, let $Q_{ik} = I(\text{patient } i \text{ eligible for process } k)$ and $\hat\pi_{ak} = n_{ak}/n_a$. Per-process influence functions on the eligible subset are embedded in the arm-level matrix with eligibility weight $Q_{ik}/\hat\pi_{ak}$; ineligible patients contribute zero.

### Data

Multivariate data are in **long format** with an `event_type` column on every row, plus `idx`, `arm`, `time`, and `status`. Set `n_event_types > 1` in `GenData` to simulate such data.

#### Risk sets and terminal events

A subject is at risk for process $k$ if they have at least one row with non-missing `event_type = k` (recurrent or typed terminal). Subjects with no type-$k$ rows are excluded from $\hat\Delta_k(\tau)$. Terminal rows are interpreted as follows:

| `status` | `event_type` | Effect |
|----------|--------------|--------|
| 0 or 2   | missing      | Ends follow-up for all processes that rely on it |
| 0 or 2   | $k$          | Ends follow-up for process $k$ only |

A global terminal row does not place a subject in the risk set for processes without typed rows. To include a subject at risk for $k$ with zero recurrent type-$k$ events, add a typed censoring or death row with `event_type = k`.

For each event type $k$ for which a subject is at risk, follow-up must end with a single observation-terminating event. By default (`cens_after_last = TRUE`), a censoring event is added after the last recurrent event for each remaining at-risk event type when no typed terminal row is present. Set `cens_after_last = FALSE` to leave such patients at risk indefinitely (with a warning).

### Analyses

#### Multivariate AUCs

```{r}
covariates <- data.frame(
  arm = c(rep(1, 100), rep(0, 100))
)
mv_data <- MCC::GenData(
  beta_event = c(log(0.8)),
  covariates = covariates,
  n_event_types = 2L,
  base_event_rate = c(1.0, 0.8),
  death_frailty_variance = 0.2,
  event_frailty_variance = 0.2,
  tau = 4
)
head(mv_data)
```

```{r}
mv_aucs <- MCC::CompareMvAUCs(
  mv_data,
  tau = 4,
  reps = 200
)
show(mv_aucs)
```

By default, inference is asymptotic (`reps = NULL`); setting `reps` adds subject-level bootstrap CIs and p-values per event type. Optional `covars` invoke multivariate augmentation. A weighted scalar contrast is available via `process_weights`:

```{r}
mv_weighted <- MCC::CompareMvAUCs(
  mv_data,
  tau = 4,
  process_weights = c(1, 1)
)
print(mv_weighted@Weighted, row.names = FALSE)
```

#### Plotting

```{r, fig.width=7, fig.height=4}
q_mv <- MCC::PlotMvMCFs(
  mv_data,
  tau = 4,
  x_breaks = seq(0, 4, by = 1),
  event_type_labs = c("Hospitalization", "Heart failure")
)
q_mv
```

See the [plotting vignette](https://github.com/zrmacc/MCC/blob/master/vignettes/plotting.pdf) for additional multivariate plotting options.
