---
title: "Plotting the Mean Cumulative Function"
author: "Zachary McCaw"
date: "`r Sys.Date()`"
output: 
  pdf_document
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Plotting}
  %\usepackage[UTF-8]{inputenc}
--- 


```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.align = "center")
suppressPackageStartupMessages({
  library(dplyr)
  library(MCC)
})
```

# Plotting the Mean Cumulative Function

## One sample setting

### Simulate data for plotting
```{r}
data <- MCC::GenData(n = 100)
```

\newpage
### Plot the MCF with NARs
```{r, fig.align='center'}
q_main <- MCC::PlotMCFs(
  data = data,
  tau = 4,
  x_breaks = seq(from = 0, to = 4),
  color_lab = "Single Arm"
)
show(q_main)
```

\newpage
### Plot area under the MCF with NARs
```{r, fig.align='center'}
q_main <- MCC::PlotMCFs(
  data = data,
  tau = 4,
  x_breaks = seq(from = 0, to = 4),
  color_lab = "Single Arm",
  show_auc = TRUE
)
show(q_main)
```

\newpage
## Two sample setting

### Simulate data for plotting
```{r}
covariates <- data.frame(
  arm = c(rep(1, 100), rep(0, 100)),
  covar = stats::rnorm(200)
)
data <- MCC::GenData(
  beta_event = c(log(0.8), log(1.2)),
  covariates = covariates,
  frailty_variance = 0.2,
  tau = 4
)
```

\newpage
### Plot the MCFs with NARs
```{r, fig.align='center'}
q_main <- MCC::PlotMCFs(
  data = data,
  tau = 4,
  x_breaks = seq(from = 0, to = 4),
  color_labs = c("Control", "Treatment")
)
show(q_main)
```

\newpage
### Plot one arm with area under the MCF
```{r, fig.align='center'}
q_main <- MCC::PlotMCFs(
  data = data,
  which_arm = 0,
  tau = 4,
  show_nar = FALSE,
  show_auc = TRUE,
  color_lab = "Control"
)
show(q_main)
```

\newpage
## Multivariate recurrent events

### Simulate multivariate data for plotting
```{r}
mv_covariates <- data.frame(
  arm = c(rep(1, 100), rep(0, 100))
)
mv_plot_data <- MCC::GenData(
  beta_event = c(log(0.8)),
  covariates = mv_covariates,
  n_event_types = 2L,
  death_frailty_variance = 0.2,
  event_frailty_variance = 0.2,
  tau = 4
)
```

\newpage
### Plot multivariate MCFs with NARs
```{r, fig.align='center'}
q_mv_main <- MCC::PlotMvMCFs(
  data = mv_plot_data,
  tau = 4,
  x_breaks = seq(from = 0, to = 4),
  color_labs = c("Control", "Treatment"),
  event_type_labs = c("Hospitalization", "Heart failure")
)
show(q_mv_main)
```

\newpage
### Plot one arm with area under the MCF
```{r, fig.align='center'}
q_mv_auc <- MCC::PlotMvMCFs(
  data = mv_plot_data,
  which_arm = 0,
  tau = 4,
  show_nar = FALSE,
  show_auc = TRUE,
  color_lab = "Control",
  event_type_labs = c("Hospitalization", "Heart failure")
)
show(q_mv_auc)
```
